From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-30 Pi/paper/discussion.tex

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work, as in~\cref{fig:table-summary}. In the
remainder of this section, we discuss some broader related work.

\begin{figure}
    \begin{center}
        \begin{tabular}{|c|c|c|c|}
            \hline
            \rowcolor{lightgray} $\PiLang$     & $\bigsqcup_n \mathcal{B}\Sn $ & $\BFin$     & $\UFin$ \tabularnewline
            \hline
            types         & Natural numbers                & Finite sets & 0-cells \tabularnewline
            \hline
            1-combinators & Generators of $\Sn$            & Bijections  & 1-paths \tabularnewline
            \hline
            2-combinators & Relations of $\Sn$             & Homotopies  & 2-paths \tabularnewline
            \hline
        \end{tabular}
    \end{center}
    \caption{Summary of the established correspondences}
    \label{fig:table-summary}
\end{figure}

@@

%%%%%%%%%%%%%%
\paragraph{Reversible Programming Languages} The practice of programming languages is replete with \emph{ad hoc}
instances of reversible computations: database transactions, mechanisms for data provenance, checkpoints, stack and
exception traces, logs, backups, rollback recoveries, version control systems, reverse engineering, software
transactional memories, continuations, backtracking search, and multiple-level undo features in commercial
applications. In the early nineties, \citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
treatment of reversibility. But intensive research in full-fledged reversible models of computations and reversible
programming languages was only sparked by the discovery of deep connections between physics and
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275,Hey:1999:FCE:304763,fredkin1982conservative},
and by the potential for efficient quantum computation~\cite{springerlink:10.1007/BF02650179}. The early developments of
reversible programming languages started with a conventional programming language, e.g., an extended $\lambda$-calculus,
and either (i) extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or (ii) imposed constraints on the control
flow constructs to make them reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.  More modern approaches recognize that
reversible programming languages require a fresh approach and should be designed from first principles without the
detour via conventional irreversible
languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}. The version of
$\PiLang$ studied in this paper is restricted to finite types and terminating total computations. It would be
interesting to understand which versions of free monoidal structures correspond to extensions of $\Pi$ with recursive
types~\cite{jamesInformationEffects2012,rc2011} and negative/fractional
types~\cite{chenComputationalInterpretationCompact2021}.

%%%%%%%%%%%%%%
\paragraph{Permutations} Finding formal systems for expressing various flavors of computable functions has been a major
focus of logic and computer science since its inception. Permutations, being at the core of reversible computing, are an
interesting class of functions, for which there are few formal systems. We develop such a system bringing in all the
associated benefits of syntactic calculi, notably, their calculational flavor. Instead of comparing two reversible
programs by extensional equality of the underlying bijective functions, a calculus offers more nuanced techniques that
can enforce additional intensional constraints on the desired equality relation.

>>>>

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible computing based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014}, and the Lambek correspondence was
suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels}. In this work, we have established this
correspondence, as summarised in~\cref{fig:table-summary}. In the remainder of this section, we discuss some broader
related work.

\begin{figure}
    \begin{center}
        \begin{tabular}{|c|c|c|c|}
            \hline
            \rowcolor{lightgray} $\PiLang$ & $\bigsqcup_n \mathcal{B}\Sn $ & $\BFin$     & $\UFin$ \tabularnewline
            \hline
            Types                          & Natural numbers               & Finite sets & 0-cells \tabularnewline
            \hline
            1-combinators                  & Generators of $\Sn$           & Bijections  & 1-paths \tabularnewline
            \hline
            2-combinators                  & Relations of $\Sn$            & Homotopies  & 2-paths \tabularnewline
            \hline
        \end{tabular}
    \end{center}
    \caption{Summary of the established correspondences}
    \label{fig:table-summary}
\end{figure}

@@

%%%%%%%%%%%%%%
\paragraph{Reversible Programming Languages} The practice of programming languages is replete with \emph{ad hoc}
instances of reversible computations: database transactions, mechanisms for data provenance, checkpoints, stack and
exception traces, logs, backups, rollback recoveries, version control systems, reverse engineering, software
transactional memories, continuations, backtracking search, and multiple-level undo features in commercial applications.
In the early nineties, \citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class, treatment of
reversibility. But intensive research in full-fledged reversible models of computations and reversible programming
languages was only sparked by the discovery of deep connections between physics and
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275,Hey:1999:FCE:304763,fredkin1982conservative},
and by the potential for efficient quantum computation~\cite{springerlink:10.1007/BF02650179}. The early developments of
reversible programming languages started with a conventional programming language, e.g., an extended $\lambda$-calculus,
and either (i) extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or (ii) imposed constraints on the control
flow constructs to make them reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.  More modern approaches recognize that
reversible programming languages require a fresh approach and should be designed from first principles without the
detour via conventional irreversible
languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}. The version of
$\PiLang$ studied in this paper is restricted to finite types and terminating total computations. It would be
interesting to understand which versions of free monoidal structures correspond to extensions of $\Pi$ with recursive
types~\cite{jamesInformationEffects2012,rc2011}, or negative and fractional
types~\cite*{chenFractionalTypes2020,chenComputationalInterpretationCompact2021}.

%%%%%%%%%%%%%%
\paragraph{Permutations} Finding formal systems for expressing various flavors of computable functions has been a major
focus of logic and computer science since its inception. Permutations, being at the core of reversible computing, are an
interesting class of functions, for which there are few formal systems. The system we have developed brings in all the
associated benefits of syntactic calculi, notably, their calculational flavor. Instead of comparing two reversible
programs by extensional equality of the underlying bijective functions, a calculus offers more nuanced techniques that
can enforce additional intensional constraints on the desired equality relation.
